---
title: "Frozen Civil Conflicts"
subtitle: "The Role of External State Support"
author: J.L.A. Krusell
date: "`r Sys.Date()`"
linestretch: 1.25
bibliography: library.bib
format: pdf
execute:
    echo: false
    warning: false
abstract: |
    Frozen conflicts are among some of the most pernicious, difficult to
    resolve conflicts. Surprisingly, little scholarly attention has been
    paid to the factors influencing their emergence. This paper addresses
    one possible dimension, namely the effect of external state support to
    conflict actors on the probability of an ongoing civil conflict
    becoming frozen. I use an original dataset coding frozen conflicts as
    an outcome among UCDP civil conflict episodes from 1975 to 2019 in
    order to construct a series of bayesian, hierarchical probit models
    motivated by an application of bargaining theory to the domain of
    frozen conflicts. Results indicate that directed support to rebels in
    the absence of similar support to the government side increases the
    probability that a civil conflict episode terminates as a frozen
    conflict. All other treatment conditions --- specifically, directed
    support provided only to the government and directed supported given
    to both sides --- are found to be inconsequential. Taken together,
    these findings help provide a more nuanced understanding of conflict
    termination beyond the common assumption that conflicts either
    conclude through a negotiated settlement or a one-sided military
    victory. Furthermore, this paper also helps shed light on the
    conditions under which actors may prefer a sub-optimal solution,
    allowing a conflict to transition into a state of unstable peace with
    an ever present threat of violent re-escalation, rather than a
    conclusive termination.
---

[//]: # Setup R environment
```{r setup, include = F}
library(dplyr)
library(flextable)
library(ggplot2)
library(ggthemes)
library(ggdist)
library(modelsummary)
library(knitr)
library(patchwork)
library(tidyr)

# Project specific utility package
library(fc.utils)

options(knitr.kable.NA = ".", knitr.kable.format = "latex",
        knitr.kable.linesep = "")

theme_set(theme_tufte(base_family = "TeX Gyre Pagella", base_size = 10))

design <- readRDS("./data/models/design.rds")
```

# Introduction

**Research question: to what extent does external state support to conflict actors affect the probability that a civil conflict becomes frozen?**

**Motivation**

Contribution: (*conceptual* and *empirical*)

- *Original Data*, conceptual/descriptive claim about prevalence of frozen conflicts
- Novel empirical analysis, provides a more nuanced analysis of *conflict termination* and contributes to our understanding of how external support can shape the dynamics of conflicts

The paper will argue that there exists both a theoretical and empirical gap. Studies on conflict termination have generally considered military victory and peace agreements as the only outcome. I further disaggregate the notion of termination to cover also the situation where conflicts are not conclusively termination and instead become "frozen."

In doing so, I show that frozen conflicts are more prevalent than thought and conceptually can and should cover a greater range of cases than just /de facto/ states. I also distinguish the concept of frozen conflicts from the related notions of stalemates and intractable rivalries.

I provide a novel quantitative analysis on the likelihood of a conflict becoming frozen. Specifically, I examine the effect of support by external state actors to belligerents in the course of a civil conflict episode. This will contribute both to our understanding of different types of conflict termination as well as the ways in which external support can shape the dynamics of conflicts.

- Potential shortcoming: what I am not doing is taking into account the preferences of external supporters and motivation for offering support. Nor am I distinguishing between different external support types.

**Why should we care about frozen conflicts?**

# Previous Literature

There exists extensive work on the impact of external third party interventions on civil conflict outcomes [@regan_third-party_2002; @balch-lindsay_third-party_2008; @gent_going_2008]. This literature has emphasised the importance of targeted interventions, either military or economic, in favor of the opposition as increasing the probability of either a negotiated settlement or rebel military victory. The opposition is generally assumed to be the weaker side at the onset of fighting such that biased interventions in favor of the opposition can result in a much larger shift in the balance of power, allowing the opposition to credibly challenge the government. Subsequent research has further disaggregated the types of support that are offered to conflict belligerents, arguing in contrast that fungible support in particular actually decreases the likelihood of a negotiated settlement and increases conflict duration by contributing to information uncertainty over the true strength of the opposition [@sawyer_role_2017; @karlen_escalate_2023]. Furthermore, external support has been found to also increase the likelihood of conflict recurrence by possibly creating an expectation of renewed assistance which may cause actors to be less risk averse and willing to militarily relitigate past conflicts [@karlen_legacy_2017].

The common thread across these works is a general assumption that
conflicts terminate either with a one-sided military victory or a negotiated settlement. Frozen conflicts, however, represent a qualitatively different type of conflict termination. Unlike a one-sided military victory, there is no clear victor when a conflict freezes, rather the conflict stalemates with both sides either unwilling or unable to continue large-scale hostilities. And, compared to a negotiated settlement, although frozen conflicts can be preceded by a ceasefire or partial settlement, they crucially do not encompass comprehensive peace agreements which would otherwise conclusively terminate a conflict. Thus, frozen conflicts represent a unique outcome; one that has previously not been examined, especially in regards to international dynamics such as external state support.

# Defining Frozen Conflicts

I follow the definition of frozen conflicts offered by @smetana_between_2019 as a "protracted, post-war conflict process, characterized by the absence of stable peace...[where] core issues between the opposing sides remain unresolved, the dispute is in the forefront of mutual relations, and there is a looming threat of the renewal of violence."  Accompanying this definition, @smetana_between_2019 lay forth three criteria determining whether a post-conflict period qualifies as a frozen conflict. First, there needs to be an initial period of active conflict followed by a cessation of large-scale violent hostilities. Isolated, sporadic incidences of violence may continue to occur; however, conflict actors are either unwilling or unable to launch major military offensives. Although not necessary, this may be characterized by the establishment of a general ceasefire and/or presence of external peacekeeping troops.  Second, the core issue underlying the conflict needs to remain salient post-conflict. This entails actors maintaining their claims within a dispute. And, finally, there must be an ongoing threat of violent re-escalation. In other words, post-conflict relations should be characterized by mutual hostility and an unwillingness to abandon violence as a political strategy.

These criteria distinguish frozen conflicts from related concepts such as protracted stalemates or enduring rivalries. In the case of the former, stalemates during a conflict can potentially endure for extended periods of time and offer conditions similar to those of a frozen conflict. However, they do not necessitate a cessation of active fighting. Stalemates can still occur during high intensity periods of a conflict where relatively few gains are made by either side. In the case of the latter, enduring rivalries are characterized in part by repeated onsets of militarized interstate disputes [@Goertz1992; @Goertz1993]; however, this might not entail the actual outbreak of violence. More importantly, enduring rivalries were originally conceptualized in terms of inter-state relations, while the definition of frozen conflicts here can be extended to the intra-state case as well.

# Theory

Why do external states intervene in civil conflicts? To further their national interests. Interveners face constraints, strategic decisions over limited resources, as well as from other interveners (ex: Anderson 2019)

**Fundamental question**: why would actors accept a frozen conflict over a comprehensive peace agreement?

When a conflict freezes, actors elide the ongoing costs of fighting; however, there remains an ever-present threat of violent re-escalation. Thus, actors continue to bear the cost of militarization, lest there is an unfavourable shift in the balance of power enabling the opposing side to re-launch the conflict.[^recurrence] The question then is, why bear the cost of an ongoing frozen conflict, as well as the uncertainty of re-escalation which ensures that war spoils remain transient? A comprehensive peace agreement, especially with international guarantees, solidifies the current status quo and allows actors to translate war spoils to permanent gains.

[^recurrence]: Conflict recurrence can still be a threat even in the case of a conclusive termination, *i.e.*, one-sided military victory or comprehensive peace agreement. However, the uncertainty around future violence in a frozen conflict is higher since there is neither a *de jure* agreement settling the underlying issue or the payoff of a total victory where the victor can dictate the post-conflict status quo.

Three possible reasons for why actors may prefer to let a civil conflict freeze:

1. **Issue Indivisibility**: Issue indivisibility is often times downplayed in bargaining theory --- it is often assumed that there exists a possible division of contested goods even in the most intractable conflicts --- however, it could play an important role in frozen conflicts. Particularly from the perspective of the government when a negotiated settlement may involve the loss of territory to a rebel group. There exists strong norms around territorial integrity as well as domestic pressure which make it difficult for states to acquiesce to secession. Furthermore, in highly fractionalized states, the government has a strong incentive to refuse a negotiated settlement in order to deter future sovereignty claims by other groups.

2. **Credible Committments**: For separatist conflicts it is often difficult for the government to credibly commit to an agreement granting greater autonomy to a separatist region. For the rebels, *de facto* independence as a half-way measure, achieved through a frozen conflict, can be the best possible outcome given that the international community tends to be fairly conservative regarding sovereignty claims. Even in the case of conflicts where the core contestation is over control of the central government, there exists a significant amount of literature pointing to the difficulties of guaranteeing the integrity of the incumbent, especially in authoritarian contexts were there is a fear of retribution.

3. **Time Horizons**: Probably the most significant mechanism concerns actors with forward-looking time horizons. Frozen conflicts, much like ceasefires which are often times involved in creating the conditions for a frozen conflict, can be a delaying tactic if an actor believes that future shifts in the distribution of power will be favourable to their cause. This can be especially important for separatist conflicts where maintaining sovereignty claims over a territory allows for an easy pretext for future escalations. International norms against wars of territorial conquest make relitigating a separatist war where the rebels have achieved a *de jure* independent state significantly more difficult, as compared to a conflict that remains a civil war within the eyes of the international community.

How does external support affect this calculus?

1. Balance of power. This mirrors the theoretical arguments made in the previous literature on external support and conflict outcomes. Essentially, support to a belligerent increases their fighting capacity, and especially in the case of a rebel group that is severely outmatched by a government, increases the likelihood of a military stalemate. This does not mean that the conflict will automatically become frozen, but it is an important first step. Furthermore, the involvement of third party actors can complicate peace negotiations since interests may not also be aligned between the donor and recipient, or even between donor states.

2. External support may also create an expectation that support will continue in the future such that an actor can offset the costs of an ongoing frozen conflict. For example, the post-Soviet separatist states are entirely reliant on Russian support, which allows them to maintain a state of militarization that deters re-escalation by the incumbent state.

# Empirics

To empirically assess whether external state support for conflict actors affects the probability of a civil conflict episode terminating as a frozen conflict a series of bayesian probit models were run using original data on frozen civil conflicts. A mixed identification strategy was employed by both conditioning on relevant covariates and instrumenting external support with UN roll call votes.

In what follows, I will first present the model specification and results from a probit regression predicting the onset of a frozen conflict. I will then decrease the dependent and independent variables, including operationalizations and data sources.

## Skeleton

Bayesian probit model

Let $y_i \in \{ 0, 1 \}$ be abinary indicator

- Unit of analysis - civil conflict episodes

  - Contiguous years of active fighting according to the UCDP definition

  - Justification --- not interested in duration or predicting
    __when__ a conflict terminates

- X a matrix of control covariates

- T a binary treatment matrix consisting of 3 variables:

    1. External state support for rebels
    2. External state support for the government
    3. Interaction between previous two indicators to capture the
       differential effect of both sides receiving external state
       support.

- Let $y^\star$ be a latent dependent variable such that ... regression equation ... and $P(y_i = 1) = \Phi(...)$ where $\Phi(\cdot)$ is the standard normal CDF.

Then $\delta_{k_i}$ is the vector of group varying treatment effects. $\gamma_{c_i}$ is a country specific intercept to capture heterogeneity

I place weakly information priors with hierarchical priors on $\delta$ and $\gamma$ in order to partially pool the estimators.

... prior equations ...

## Probit Model

I model the onset of a frozen civil conflict using a bayesian probit regression where the unit analysis is civil conflict episodes, *i.e.*, contiguous years of active fighting within a conflict. The outcome variable, $y \in \{ 1, 2 \}$, is a binary indicator denoting the


TODO: eventually I would like to move towards a dyadic analysis, rather than conflict level

- Unit of analysis: conflict episodes, ie consecutive years of active fighting (above UCDP threshold of 25 battle deaths)
- Binary outcome, binary treatment, while conditioning on a number of relevant covariates

\begin{align*}
y_i^\star & = \alpha + X_i' \beta + T_i' \delta + \epsilon_i, \; \epsilon_i \sim \text{N}(0, \sigma^2) \\
y_i & = \begin{cases}
1 & \text{if } y_i^\star > 0 \\
0 & \text{otherwise}
\end{cases}
\end{align*}

- Estimate with Stan, weakly informative priors, $\delta \sim \text{Normal}(1, 1)$

## Data

### Measuring Frozen Conflicts

In order to measure the onset of frozen civil conflicts I followed the previous three criteria and coded an original dataset using dyadic intra-state conflict episodes drawn from the UCDP/PRIO Armed Conflict Dataset [@gleditsch_armed_2002; @davies_organized_2023] defined as contiguous years with at least 25 battle-related deaths in a violent contest over a government or territory between a state and non-state actor. To address post-conflict issue saliency, I considered an episode to be unresolved if it terminated without either a comprehensive peace treaty addressing the core underlying issues or a one-sided military victory for either the government or rebel group according to the UCDP Peace Agreement and Conflict Termination Datasets [@pettersson_organized_2019; @kreutz_how_2010]. I also filtered out episodes where one or more actors ceased to exist within five years of the conflict termination and where the non-state actor renounces violences. This was done using a manual coding process with a mix of academic and news media sources for different conflicts in order to capture instances where, for example, rebel groups transition into legitimate political parties.

```{r}
dataset <- readRDS("./data/model_data.rds")
n_frozen <- filter(dataset, frozen == 1) |> nrow()
n_terr_frozen <- filter(dataset, frozen == 1, incompatibility == 1) |> nrow()
```

The resulting dataset consists of `r n_frozen` frozen conflicts out
of `r nrow(dataset)` intra-state episodes, comprising approximately
`r signif(n_frozen / nrow(dataset), 2) * 100` percent of all observations for the period 1975 to 2019. Of the frozen conflicts, `r n_terr_frozen` are conflicts over territory. (TODO: show world map and time plot)

```{r}
frozen <- filter(dataset, frozen == 1) |>
    mutate(country = sub("Government of ", "", side_a),
           country = case_when(country == "Russia (Soviet Union)" ~ "Russia",
                               country == "United Kingdom" ~ "UK",
                               country == "Congo" ~ "Republic of Congo",
                               country == "Myanmar (Burma)" ~ "Myanmar",
                               country == "Bosnia-Herzegovina" ~ "Bosnia and Herzegovina",
                               country == "Serbia (Yugoslavia)" ~ "Serbia",
                               country == "DR Congo (Zaire)" ~ "Democratic Republic of the Congo",
                               T ~ country))

sub.df <- group_by(frozen, country) |> count(frozen)

world <- map_data("world") |>
    left_join(sub.df, by = c("region" = "country")) |>
    mutate(n = ifelse(is.na(n), 0, 1) )

ggplot(world, aes(long, lat, map_id = region, fill= n, group = group)) +
    geom_map(map = world) +
    theme(legend.position = "none") +
    scale_color_solarized()
```

```{r}
filter(dataset, frozen == 1) |>
ggplot(aes(year)) +
    geom_bar() +
    xlab("Year") +
    ylab("Frozen Civil Conflict Onsets")
```

Compared to the Frozen Conflict Dataset by @klosek_frozen_2021,

Compared to FCD --- significant more cases
  - More expansive operationalization
    - Not only looking at de facto states
    - Captures frozen moments in civil wars over government, ex Angola

  - Timing of Frozen conflicts
    - FCD last year in which >= 500 BRD
    - My datasets follows UCDP last year >25
      - Captures lower level conflicts

Potential Weaknesses
  - Low-level long running insurgencies may not be "salient"
  - Termination year inexact
  - Potentially capturing shorter stalemates

### Treatments

UCDP external support dataset, binary variable for whether an external
state actor offered either fungible or non-fungible support at least
once to either the rebels or government side in a conflict episode.

- Include interaction term, three treatment conditions:

1. No support (control group)
2. Gov. support only
3. Reb. support only
4. Both support

```{r}
conditions <- dataset |>
    mutate(treatment = case_when(ext_sup_s_rebel_bin == 1 & ext_sup_s_state_bin == 1 ~ "Ext. Sup. to Both",
                                 ext_sup_s_rebel_bin == 1 & ext_sup_s_state_bin == 0 ~ "Ext. Sup. Rebel Only",
                                 ext_sup_s_rebel_bin == 0 & ext_sup_s_state_bin == 1 ~ "Ext. Sup. Gov Only",
                                 ext_sup_s_rebel_bin == 0 & ext_sup_s_state_bin == 0 ~ "Ext. Sup. to Neither"))

m <- table(conditions$frozen, conditions$treatment)
rownames(m) <- c("Not Frozen", "Frozen")
knitr::kable(m)
```

### Control Variables

- Episode Intensity
- Episode Duration
- Incompatibility? (ie separatist conflict over territory)
- Recurrent episode
- Right censored

Descriptive statistics split between positive and negative outcomes

### Descriptive Statistics

## Observational Models

### Results

Figure 1 shows the parameter estimates for $\delta$.

```{r}
#| fig.cap: Posterior estimates for \delta from a single variate probit model
main_model <- filter(design, outcome == "frozen", treatment == "bin", episodes == "all",
                     censored == "with_censored") |> pull(id)

fit <- sprintf("posteriors/data_%s/fit.rds", main_model) |>  readRDS()

delta <- fit$draws("delta", format = "df") |>
    select(matches("delta")) |>
    pivot_longer(everything(), names_to = "variable")

ll <- extract_digits(delta$variable)
delta$variable <- sprintf("delta[%s]", sapply(ll, `[`, 1))
delta$group <- ifelse(sapply(ll, `[`, 2) == 1, "Government Incompatibility",
                      "Territorial Incompatibility")

ggplot(delta, aes(x = value, y = variable, fill = variable)) +
    stat_halfeye(alpha = 0.8, show.legend = F) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.2) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
    scale_y_discrete(labels = c("delta[1]" = "Ext. Sup. to Gov",
                                "delta[2]" = "Ext. Sup. to Rebel",
                                "delta[3]" = "Interaction")) +
    facet_grid(. ~ group) +
    scale_fill_solarized()
```

Figure 2 shows the average marginal effects, *i.e.* the change in the posterior predicted probability between each treatment condition and the control condition averaged across the sample.

\begin{equation}
\text{AME}_j = \sum_{i=1}^N \Phi(\alpha + X_i' \beta + \delta_i) - \Phi(\alpha + X_i' \beta)
\end{equation}


```{r}
#| fig.cap: Average marginal effects for the three treatment conditions on the predicted probability of a frozen conflict onset.[^ame]
ame <- fit$draws("ame", format = "df") |>
    select(matches("ame")) |>
    pivot_longer(everything(), names_to = "variable")

ll <- extract_digits(ame$variable)
ame$variable <- sprintf("ame[%s]", sapply(ll, `[`, 1))
ame$group <- ifelse(sapply(ll, `[`, 2) == 1, "Government Incompatibility",
                    "Territorial Incompatibility")

ggplot(ame, aes(x = value, y = variable, fill = variable)) +
    stat_eye(alpha = 0.8, show.legend = F) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.2) +
    theme(axis.title.x = element_blank()) +
    labs(x = parse(text = "bold('AME '*Delta)")) +
    scale_y_discrete(labels = c("ame[1]" = "Ext. Sup. Gov Only",
                                "ame[2]" = "Ext. Sup. Rebel Only",
                                "ame[3]" = "Ext. Sup. to Both")) +
    coord_flip() +
    facet_grid(. ~ group) +
    scale_fill_solarized()
```

[^ame]: Hello World

- Parameter Plot
- Density of predicted probabilities for true success/failure
- Marginal effects
- Confusion Matrix + AUR/PRC plots

#### Robustness Checks

- External support only within the last five years of conflict
- Frozen conflicts with a duration of only 1 year
- Exclude low intensity conflicts?

Appendix:

- Proportion of episode years with external support
- Nr of external supporters, maybe weighted by cinc?
- Territorial conflicts only
- FCD data

## IV Model

\newpage

# References

::: {#refs}
:::

\newpage

# Appendix

## List of Frozen Conflicts

```{r}
filter(dataset, frozen == 1) |>
    arrange(side_a, year) |>
    select(Government = side_a, `Rebel Group(s)` = side_b, Year = year) |>
    mutate(Government = sub("Government of", "", Government) |> trimws()) |>
    kable(longtable = T)
```

## Simulation Based Calibration

## Robustness Checks

Different combinations of the following...

- Aggregation of the treatments: binary treatment, binary treatment in the last 5 years of the conflict episode, and proportional treatment. The latter is calculated as the proportion of conflict-years in an episode where one side received external support

- All conflict episodes and only high intensity episodes --- try to distinguish between low-level insurgencies and actual civil wars

- Normal definition for frozen conflicts and a strict definition that requires a duration of at least 2 years in order to exclude temporary one-year pauses in fighting that may be more the result of UCDP's coding decision rather than representing actual frozen conflict.


```{R}
pzero <- function(x) mean(x > 0)

models <- list.dirs("posteriors", recursive = F)
stopifnot(length(models) != 0)

fits <- lapply(models, \(dir) file.path(dir, "fit.rds") |> readRDS())
names(fits) <- basename(models)

model_ames <- lapply(fits, function(fit) {
    fit$summary("ame", pzero, posterior::quantile2, median)
}) |> bind_rows()

ll <- extract_digits(model_ames$variable)
ame_keys <- sprintf("ame[%s]", sapply(ll, `[`, 1))

model_ames$variable <- case_when(ame_keys == "ame[1]" ~ "Ext. Sup. Gov Only",
                                 ame_keys == "ame[2]" ~ "Ext. Sup. Rebel Only",
                                 ame_keys == "ame[3]" ~ "Ext. Sup. to Both")
model_ames$group <- ifelse(sapply(ll, `[`, 2) == 1, "Government Incompatibility",
                           "Territorial Incompatibility")

ggplot(model_ames, aes(pzero, median)) +
    geom_point(alpha = 0.2) +
    geom_errorbar(aes(ymin = q5, ymax = q95), alpha = 0.25) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.2) +
    facet_grid(variable ~ group) +
    labs(y = parse(text = "tilde(Delta)"), x = parse(text = "P(tilde(Delta) > 0)")) +
    theme(legend.position = "none")

```
