---
title: "From Active Hostilities to Frozen Stalemates: Assessing the Role of External Support in Civil Conflict Dynamics"
author: J.L.A. Krusell
date: "`r Sys.Date()`"
linestretch: 1.25
bibliography: library.bib
format: pdf
execute:
    echo: false
    warning: false
---

[//]: # Setup R environment
```{r setup, include = F}
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggdist)
library(modelsummary)
library(kableExtra)
library(knitr)
library(patchwork)
library(tidyr)

# Project specific utility package
library(fc.utils)

options(knitr.kable.NA = ".", knitr.kable.format = "latex",
        knitr.kable.linesep = "")

theme_set(theme_tufte(base_family = "TeX Gyre Pagella", base_size = 10))
```

**Research question: to what extent does external state support to conflict actors affect the probability that a civil conflict becomes frozen?**

# Empirics

To empirically assess whether external state support for conflict
actors affects the probability of a civil conflict episode terminating
as a frozen conflict a series of bayesian probit models were run
using original data on frozen civil conflicts. A mixed identification strategy was employed by both conditioning on relevant covariates and instrumenting external support with UN roll call votes.

In what follows, I will first describe the dependent and independent variables, including operationalizations and data sources. I will then present the model specification and results from a conventional probit regression followed by the instrumental variable analysis.

## Data

### Defining Frozen Conflicts

I follow the definition of frozen conflicts offered by @smetana_between_2019 as a "protracted, post-war conflict process, characterized by the absence of stable peace...[where] core issues between the opposing sides remain unresolved, the dispute is in the forefront of mutual relations, and there is a looming threat of the renewal of violence."  Accompanying this definition, @smetana_between_2019 lay forth three criteria determining whether a post-conflict period qualifies as a frozen conflict. First, there needs to be an initial period of active conflict followed by a cessation of large-scale violent hostilities. Isolated, sporadic incidences of violence may continue to occur; however, conflict actors are either unwilling or unable to launch major military offensives. Although not necessary, this may be characterized by the establishment of a general ceasefire and/or presence of external peacekeeping troops.  Second, the core issue underlying the conflict needs to remain salient post-conflict. This entails actors maintaining their claims within a dispute. And, finally, there must be an ongoing threat of violent re-escalation. In other words, post-conflict relations should be characterized by mutual hostility and an unwillingness to abandon violence as a political strategy.

These criteria distinguish frozen conflicts from related concepts such as protracted stalemates or enduring rivalries. In the case of the former, stalemates during a conflict can potentially endure for extended periods of time and offer conditions similar to those of a frozen conflict. However, they do not necessitate a cessation of active fighting. Stalemates can still occur during high intensity periods of a conflict where relatively few gains are made by either side. In the case of the latter, enduring rivalries are characterized in part by repeated onsets of militarized interstate disputes [@Goertz1992; @Goertz1993]; however, this might not entail the actual outbreak of violence. More importantly, enduring rivalries were originally conceptualized in terms of inter-state relations, while the definition of frozen conflicts here can be extended to the intra-state case as well.

### Measuring Frozen Conflicts

In order to measure the onset of frozen civil conflicts I followed the previous three criteria and coded an original dataset using dyadic intra-state conflict episodes drawn from the UCDP/PRIO Armed Conflict Dataset [@gleditsch_armed_2002; @davies_organized_2023] defined as contiguous years with at least 25 battle-related deaths in a violent contest over a government or territory between a state and non-state actor. To address post-conflict issue saliency, I considered an episode to be unresolved if it terminated without either a comprehensive peace treaty addressing the core underlying issues or a one-sided military victory for either the government or rebel group according to the UCDP Peace Agreement and Conflict Termination Datasets [@pettersson_organized_2019; @kreutz_how_2010]. I also filtered out episodes where one or more actors ceased to exist within five years of the conflict termination and where the non-state actor renounces violences. This was done using a manual coding process with a mix of academic and news media sources for different conflicts in order to capture instances where, for example, rebel groups transition into legitimate political parties.


```{r}
dataset <- readRDS("./data/model_data.rds")
n_frozen <- filter(dataset, frozen == 1) |> nrow()
n_terr_frozen <- filter(dataset, frozen == 1, incompatibility == 1) |> nrow()
```

The resulting dataset consists of `r n_frozen` frozen conflicts out
of `r nrow(dataset)` intra-state episodes, comprising approximately
`r signif(n_frozen / nrow(dataset), 2) * 100` percent of all observations for the period 1975 to 2019. Of the frozen conflicts, `r n_terr_frozen` are conflicts over territory. (TODO: show world map and time plot)

```{r}
filter(dataset, frozen == 1) |>
    ggplot(aes(year)) +
    geom_bar() +
    ylab("Frozen Civil Conflict Onsets") +
    xlab("Year")
```

Compared to the Frozen Conflict Dataset by @klosek_frozen_2021,

Compared to FCD --- significant more cases
  - More expansive operationalization
    - Not only looking at de facto states
    - Captures frozen moments in civil wars over government, ex Angola

  - Timing of Frozen conflicts
    - FCD last year in which >= 500 BRD
    - My datasets follows UCDP last year >25
      - Captures lower level conflicts

Potential Weaknesses
  - Low-level long running insurgencies may not be "salient"
  - Termination year inexact
  - Potentially capturing shorter stalemates

### Treatments

UCDP external support dataset

### Control Variables

- Conflict Intensity
- Conflict Duration
- Incompatibility?
- e_gdppc
- e_pop
- CINC

Descriptive statistics split between positive and negative outcomes

### Descriptive Statistics

## Observational Models

\begin{align*}
y_i^\star & = \alpha + X_i' \beta + T_i' \delta + \epsilon_i, \; \epsilon_i \sim \text{N}(0, \sigma^2) \\
y_i & = \begin{cases}
1 & \text{if } y_i^\star > 0 \\
0 & \text{otherwise}
\end{cases}
\end{align*}

### Results

```{r}
#| fig.cap: Posterior estimates for \delta from a single variate probit model
fit <- readRDS("./posteriors/frozen-bin-all/fit.rds")

delta <- fit$draws("delta", format = "df") |>
    select(matches("delta")) |>
    pivot_longer(everything(), names_to = "variable")

ggplot(delta, aes(x = value, y = variable, fill = variable)) +
    stat_halfeye(alpha = 0.8, show.legend = F) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.2) +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) +
    scale_y_discrete(labels = c("delta[1]" = "Ext. Sup. to Gov",
                                "delta[2]" = "Ext. Sup. to Rebel",
                                "delta[3]" = "Interaction"))
```

```{r}
#| fig.cap: Average marginal effects for the three treatment conditions on the predicted probability of a frozen conflict onset.[^ame]
ame <- fit$draws("ame", format = "df") |>
    select(matches("ame")) |>
    pivot_longer(everything(), names_to = "variable")

ggplot(ame, aes(x = value, y = variable, fill = variable)) +
    stat_eye(alpha = 0.8, show.legend = F) +
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.2) +
    theme(axis.title.x = element_blank()) +
    labs(x = parse(text = "bold('AME '*Delta)")) +
    scale_y_discrete(labels = c("ame[1]" = "Ext. Sup. Gov Only",
                                "ame[2]" = "Ext. Sup. Rebel Only",
                                "ame[3]" = "Ext. Sup. to Both")) +
    coord_flip()
```

[^ame]: Hello World

- Parameter Plot
- Density of predicted probabilities for true success/failure
- Marginal effects
- Confusion Matrix + AUR/PRC plots

#### Robustness Checks

- External support only within the last five years of conflict
- Frozen conflicts with a duration of only 1 year
- Exclude low intensity conflicts?

Appendix:

- Proportion of episode years with external support
- Nr of external supporters, maybe weighted by cinc?
- Territorial conflicts only
- FCD data

## IV Model

\newpage

# References

::: {#refs}
:::

\newpage

# Appendix

## Robust Checks

### Observational Probit Models

```{R}
models <- list.dirs("posteriors", recursive = F)
stopifnot(length(models) != 0)

fits <- lapply(models, \(dir) file.path(dir, "fit.rds") |> readRDS())
names(fits) <- basename(models)
```

```{R}
#| fig-height: 7
summary.df <- lapply(fits, function(fit) {
    post <- post_summarise(fit, "delta", rename = T, model_info = T,
                           probs = c(0.025, 0.17, 0.5, 0.83, 0.975)) |>
        rename(median = `50%`,
               lower95 = `2.5%`,
               upper95 = `97.5%`,
               lower66 = `17%`,
               upper66 = `83%`)
}) |> bind_rows()

summary.df$variable <- sub("_(prop|bin.*)$", "", summary.df$variable)

ggplot(summary.df, aes(treatment_aggregation, median, color = outcome)) +
    geom_point(size = 1.5, position = position_dodge(width = 0.25)) +
    geom_errorbar(aes(ymin = lower95, ymax = upper95), position = position_dodge(0.25), width = 0) +
    geom_errorbar(aes(ymin = lower66, ymax = upper66), position = position_dodge(0.25), linewidth = 1.25, width = 0) +
    geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.2) +
    facet_grid(variable ~ episodes)

```

```{R}
models <- list.dirs("posteriors", recursive = F)
stopifnot(length(models) != 0)

fits <- lapply(models, \(dir) file.path(dir, "fit.rds") |> readRDS())

frozen <- fits[grepl("/frozen", models)]
strict_frozen <- fits[grepl("/strict_frozen", models)]

modelsummary(frozen, pars = c("alpha", "delta", "beta"),
             statistic = "conf.int", title = "Frozen Conflicts") |>
    add_header_above(c(" " = 1, "All Episodes" = 3, "High Intensity Only" = 3)) |>
    kable_styling(font_size = 8, full_width = F) |>
    landscape()
```

```{R}
modelsummary(strict_frozen, pars = c("alpha", "delta", "beta"),
             statistic = "conf.int", title = "Frozen Conflicts Lasting >2 Years") |>
    add_header_above(c(" " = 1, "All Episodes" = 3, "High Intensity Only" = 3)) |>
    kable_styling(font_size = 8, full_width = F) |>
    landscape()
```
